Do researchers preferentially collaborate with colleagues of the same gender?

### 2018-05-29 manuscript/Manuscript_Claire.Rmd
Finalise markdown

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin&sect;
- '*\textit{luke.holman@unimelb.edu.au} \vspace{5mm}'
output:
  pdf_document: default
  html_document: default
bibliography: references.bib
fontsize: 12pt
header-includes: 
  \usepackage{microtype} 
  \usepackage[all]{hypcap} 
  \usepackage{amsmath} 
  \usepackage{booktabs}
  \usepackage{caption} 
  \usepackage[labelfont=bf]{caption} 
  \usepackage{titling} 
  \pretitle{\begin{flushleft}\LARGE}
  \posttitle{\par\end{flushleft}\vskip 0.5em} 
  \preauthor{\begin{flushleft}\large \lineskip 0.5em} 
  \postauthor{\par\end{flushleft}} 
  \predate{\begin{flushleft}\large} 
  \postdate{\par\end{flushleft}}
  \usepackage{fancyhdr} 
  \pagestyle{fancy} 
  \fancyhead[LO,LE]{\textsl{\leftmark}} 
  \rhead[]{Gender and coauthorship} 
  \usepackage[noabbrev,capitalise]{cleveref}
  \usepackage{titlefoot} 
  \usepackage{amssymb} 
  \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious biases, and often feel excluded from formal and informal opportunities for research collaboration. In addition to ensuring fairness and helping to redress gender imbalance in the academic workforce, increasing women's access to collaboration could help scientific progress by drawing on more of the available human capital. Here, we test whether researchers preferentially collaborate with same-gendered colleagues, using more stringent methods and a larger dataset than in past work. Our results reaffirm that researchers co-publish with colleagues of the same gender, and show that this 'gender homophily' is slightly stronger today than it was 10 years ago. Contrary to our expectations, we found no evidence that homophily is driven mostly by senior academics, and no evidence that homophily is strongest in fields where women are in the minority. Interestingly, homophily was negatively correlated with journal impact factor, as predicted if mixed-gender teams produce better research. We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining, Women in STEM.
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{§Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

Women are substantially underrepresented in many branches of the workforce in science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011] than men. Experimental studies suggest that women's research achievements are regarded less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

@@

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to assign a gender to each author based on their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations where available. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

@@

## Calculating $\alpha$, the coefficient of homophily
Following Bergstrom et al. [-@bergstrom_2016], we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha > 0$ suggests that same-gender authors publish together more often than expected if authors assort randomly with respect to gender (homophily), $\alpha < 0$ suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha = 0$ suggests random assortment with respect to gender. Like the Wahlund effect, $\alpha$ is borrowed from population genetics; for a set of 2-author papers, it is equivalent to the coefficient of inbreeding [@wright_1949]. Mathematical work illustrates that $\alpha$ is closely related to network-based alternative methods for quantifying homophily [@wang_2016].

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl T. Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom's example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one in each of the six possible two-author combinations. For these six papers, $\alpha = -\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators. To control for the fact that the null expectation for $\alpha$ is not zero for small datasets, we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly permuting authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate (FDR) correction to each set of p-values to account for multiple testing [@Benjamini_1995].

@@

## Minimising the Wahlund effect: research discipline and time period
To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. Specifically, we only calculated $\alpha'$ for individual journals, since papers from the same journal typically focus on one or a few closely related topics. Although some journals, e.g. _PLoS ONE_, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. _PLoS ONE_ versus _PLoS Computational Biology_, gives a rough estimate of the extent to which multidisciplinarity inflates $\alpha'$.

@@

## Minimising the Wahlund effect: author country of affiliation
A Wahlund effect could arise even if one calculates $\alpha'$ for a single discipline and time period, because of variation in the gender ratio of researchers from different countries. For example, Holman et al. [-@Holman_2018] showed that PubMed-indexed authors based in Serbia are more than twice as likely to be women as are authors based in Japan. Therefore, a dataset containing a mix of papers from teams of authors based in these two countries would contain an excess of same-sex coauthorships, even if collaboration were random with respect to gender within each country.

To address this issue, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no direct, causal effect on coauthor selection. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine if and how this effect should change our inferences. For simplicity, our calculations assume there are only two career stages, though the conclusions probably generalise to a realistic, multi-tier career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). The Online Supplementary Material gives annotated R code used to derive the theoretical expectations.

## Data availability and reproducibility
The Online Supplementary Material contains R scripts used to produce all data, figures and tables. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/.

@@

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.09{\pm}0.04$, $t_{948}$ = 2.51, p = 0.012).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.07{\pm}0.02$, $t_{1988}$ = 4.48, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals, and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained at most 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). This implies that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category 'Multidisciplinary' - which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ - did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from earlier studies, are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis to only include authors with affiliations from a single country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S5-S6). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); the average difference in $\alpha'$ was 0.002 (Figure S7). These results suggest that our findings of widespread homophily in the main analysis were not driven by a Wahlund effect resulting from gender differences between countries.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/figure5.pdf)

@@

# Discussion
We found evidence that researchers publish with same-gendered coauthors more often than expected under random assortment, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite strong: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, this effect appears even more striking: for example, if the proportion of male coauthors is 80% for men and 70% for women ($\alpha' = 0.1$), the results imply that women publish with women 50% more often than men do.

@@

Some disciplines exhibit especially skewed gender ratios; e.g. most authors in physics and computer science are men, while most authors in nursing are women [@Holman_2018]. We hypothesised that these gender-imbalanced disciplines might show the strongest gender homophily, e.g. because being in the minority might increase one's drive to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board \autoref{alpha_histograms}. Moreover, $\alpha'$ was lower for journals with a male-biased author gender ratio relative to those with an even gender ratio. Interestingly, $\alpha'$ was highest in journals with a female-biased author gender ratio. One interpretation of this result is that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since there are so few journals in which most authors are women \autoref{alpha_gender_ratio}.

We also found that gender homophily was marginally stronger in 2015-2016, relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and probably more likely) explanations for this result. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment, sexism or microaggressions from men [@jagsi_2016], which is clearly very concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], then this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if this motivation applies to more women than men. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are 'compliant [in reproducing] male-dominated scientific structures' because they do not collaborate often enough with other women (coauthorships involving two women were about 30% more common than expected under the null). In contrast, we believe that everyone (not just women) should work to ensure that they are not overlooking or excluding female colleagues, or devoting less effort to collaborations with them. Men are more often well placed to help in this regard, given their overrepresentation in jobs that allow the hiring of staff.

@@

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more often than their male counterparts [@Debarre_2018]. Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events [@Isbell_2012; @Schroeder_2013; @Klein_2017], discrimination during hiring [@MossRacusin_2017; @Reuben_2014] and discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], are only some of the small disadvantages adding up to drive women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.

>>>>

---
title: "Do researchers preferentially collaborate with colleagues of the same gender?"
author:
- Luke Holman* and Claire Morandin\textsuperscript{§}
- '*\textit{luke.holman@unimelb.edu.au}'
- \textsuperscript{§}\textit{claire.morandin@helsinki.fi} \vspace{5mm}
bibliography: references.bib
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 12pt
header-includes: \usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
  \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}
  \pretitle{\begin{flushleft}\LARGE} \posttitle{\par\end{flushleft}\vskip 0.5em} \preauthor{\begin{flushleft}\large
  \lineskip 0.5em} \postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
  \postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
  \rhead[]{Gender and coauthorship} \usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
  \usepackage{amssymb} \usepackage{rotating}
csl: the-american-naturalist.csl
abstract: Evidence suggests that women in academia are hindered by conscious and unconscious
  biases, and often feel excluded from formal and informal opportunities for research
  collaboration. In addition to ensuring fairness and helping to redress gender imbalance
  in the academic workforce, increasing women's access to collaboration could help
  scientific progress by drawing on more of the available human capital. Here, we
  test whether researchers preferentially collaborate with same-gendered colleagues,
  using more stringent methods and a larger dataset than in past work. Our results
  reaffirm that researchers co-publish with colleagues of the same gender, and show
  that this 'gender homophily' is slightly stronger today than it was 10 years ago.
  Contrary to our expectations, we found no evidence that homophily is driven mostly
  by senior academics, and no evidence that homophily is strongest in fields where
  women are in the minority. Interestingly, homophily was negatively correlated with
  journal impact factor, as predicted if mixed-gender teams produce better research.
  We discuss potential causes of gender homophily in science. \vspace{5mm} \par\noindent
  \textbf{Keywords:} Gender bias, Homophily, Scientific collaboration, Text mining,
  Women in STEM. 
---

@@

\maketitle\unmarkedfntext{*School of BioSciences, The University of Melbourne, Victoria, Australia.}
\maketitle\unmarkedfntext{\textsuperscript{§}Organismal and Evolutionary Biology Research Programme,
Faculty of Biological and Environmental Sciences, University of Helsinki, Finland.}

@@

Women are highly underrepresented in many branches of science, technology, engineering, mathematics, and medicine (STEMM), and face additional challenges and inequities relative to men [e.g. @Shaw_2012; @Lariviere_2013; @West_2013; @Elsevier_report; @Holman_2018]. On average, women occupy more junior positions [@Wutte_2007; @Reuben_2014] with lower salaries [@Trower_2002; @Umbach_2007], receive less grant money [@Hosek_2005; @OER_2008], are promoted more slowly [@Zuckerman_1987; @Rosenfeld_1991; @Long_1993; @Hopkins_2013], and are allocated fewer resources [@ODorchai_2009] and less research funding [@Feldt_1986; @Stack_2004; @Lariviere_2011] than men. Experimental studies suggest that women's research achievements are regarded less favourably than identical achievements by men [@Moss_2012; @Knobloch_2013].

@@

# Methods
## The dataset
We used the dataset of PubMed author lists from Holman et al. [-@Holman_2018]. Briefly, that dataset was created by downloading every article indexed on PubMed and attempting to infer each author's gender from their given name. Each journal was assigned to one of 107 scientific disciplines, using PubMed's journal categorisations in the interests of objectivity. Because the present study focuses on co-authorship, all single-author papers were discarded. We also discarded all papers for which we could not determine the gender of every author with ${\ge}95\%$ certainty, in order to simplify the statistical analysis. To mitigate Wahlund effects caused by variation in the gender ratio of researchers over time (see below), we also discarded all papers except those that were published either 0-1 or 10-11 years before the PubMed data were collected (i.e. 20^th^ August 2016). Lastly, we excluded journals with fewer than 50 suitable papers.

@@

## Calculating $\alpha$, the coefficient of homophily
Following Bergstrom et al. [-@bergstrom_2016], we defined the coefficient of homophily as $\alpha = p - q$, where $p$ is the probability that a randomly-chosen co-author of a _male_ author is a man and $q$ is the probability that a randomly-chosen co-author of a _female_ author is a man. Therefore, $\alpha > 0$ suggests that same-gender authors publish together more often than expected under the null hypothesis that authors assort randomly with respect to gender (homophily), $\alpha < 0$ suggests that opposite-gender authors publish together more often than expected (heterophily), and $\alpha = 0$ suggests random assortment with respect to gender. Like the Wahlund effect, $\alpha$ is borrowed from population genetics; for a set of 2-author papers, it is equivalent to Wright's coefficient of inbreeding [@wright_1949]. Mathematical work illustrates that $\alpha$ is closely related to network-based alternative methods for quantifying homophily [@wang_2016].

@@

Our test assumes that the expected value of $\alpha$ is zero if authors randomly assort, but for small datasets this assumption is not always true (as pointed out by Carl Bergstrom in a blog post, http://www.eigenfactor.org/gender/assortativity/note_to_eisen.rtf). To borrow Prof. Bergstrom's example, consider a small research specialty comprising just two men and two women researchers, who have together produced six two-author papers: one in each of the six possible two-author combinations. For these six papers, $\alpha = -\frac{1}{3}$, even though same- and opposite-gendered coauthors were selected in equal proportion to their frequency in the pool of possible collaborators.

To control for the fact that the null expectation for $\alpha$ is not zero for small datasets, we devised an adjusted version of the coefficient of homophily, which we term $\alpha'$. Every time we calculated $\alpha$ for a set of papers, we also determined the expected value of $\alpha$ under the null hypothesis that authors assort randomly with respect to gender. This was accomplished by randomly permuting authors across papers 1000 times, recalculating $\alpha$, and taking the median. We then calculated $\alpha'$ by subtracting the null expectation for $\alpha$ from the observed value. We also used the null-simulated $\alpha$ values to calculate a two-tailed p-value for the observed value of $\alpha$; the p-value was defined as the proportion of null simulations for which $|\alpha_{null}| > |\alpha_{obs}|$. We applied false discovery rate (FDR) correction to each set of p-values to account for multiple testing [@Benjamini_1995].

@@

## Minimising the Wahlund effect: research discipline and time period
To minimise bias in $\alpha'$ due to the Wahlund effect, we restricted each set of papers to a single research specialty to the greatest extent allowed by our data. Specifically, we only calculated $\alpha'$ for individual journals, since papers from the same journal typically focus on closely related topics. Although some journals, e.g. _PLoS ONE_, publish research from diverse disciplines with very different author gender ratios [@Holman_2018], calculating $\alpha'$ for these highly multidisciplinary journals is still useful as a contrast. The difference in $\alpha'$ between highly multidisciplinary and more specialised journals, e.g. _PLoS ONE_ versus _PLoS Computational Biology_, gives an estimate of the extent to which multidisciplinarity inflates $\alpha'$.

@@

## Minimising the Wahlund effect: author country of affiliation
A Wahlund effect could arise even if one calculates $\alpha'$ for a single discipline and time period, because of variation in the gender ratio of researchers from different countries. For example, Holman et al. [-@Holman_2018] showed that PubMed-indexed authors based in Serbia are more than twice as likely to be women as are authors based in Japan. Therefore, a dataset containing a mix of papers from teams of authors based in these two countries would contain an excess of same-sex coauthorships, even if collaboration were random with respect to gender within each country. 
To address this issue, we also analysed every combination of journal and author country of affiliation for which we had enough data (i.e. 50 or more papers published in 2015-16). For simplicity, we restricted the dataset to only include papers for which Holman et al. [-@Holman_2018] had identified the country of affiliation for all authors on the paper, and all authors shared the same country of affiliation. Restricting the dataset in this fashion produced enough data to measure $\alpha'$ for 325 combinations of journal and country (median: 70 papers and 273 authors per combination).

@@

## Theoretical expectations for $\alpha$ when the gender ratio differs between career stages
In many STEMM subjects, the gender ratio is more skewed among established researchers relative to early-career researchers [@Shaw_2012; @Holman_2018]. We hypothesised that this skew could potentially create both Wahlund effects and 'reverse' Wahlund effects. For example, imagine that the majority of collaborations are between students and professors, and that the gender ratio differs between career stages: we will then see an excess of mixed-gender coauthorships (heterophily, $\alpha < 0$), even if gender has no direct, causal effect. Similarly, a hypothetical field in which students work only with students, and professors with professors, would have apparent gender homophily ($\alpha > 0$).

We can think of no tractable method of controlling for this issue using our dataset, which contains no information on career stage. Therefore, we instead decided to derive the theoretical expectations for $\alpha$ when there is a difference in gender ratio across career stages, in order to determine if and how this effect should alter our inferences. For simplicity, our calculations assume there are only two career stages, though we intuit that the general conclusions would also apply to a multi-tier career ladder. Under the null model that gender has no causal effect on collaboration, we calculated $\alpha$ for various combinations of the four free parameters, i.e. the gender ratios for early- and late-career researchers, and the relative frequency of collaborations between early-early, early-late, and late-late collaborations. We then used the theoretical expectations for $\alpha$ to qualify our main conclusions (see Results). The Online Supplementary Material gives annotated R code used to derive the theoretical expectations.

## Data availability and reproducibility
The Online Supplementary Material contains R scripts used to produce all results, figures and tables. The input data from Holman et al. [-@Holman_2018] is archived at https://osf.io/bt9ya/ along with the code used to obtain it.

@@

$\alpha'$ was significantly higher in the literature sample from 2015-16 relative to 2005-6, though the difference in means was small (Figure S3; Effect of the fixed factor 'Time period' in a linear mixed model of the data for all author positions: Cohen's $d$ = $0.091{\pm}0.04$, $t_{953}$ = 2.42, p = 0.016).

When comparing pairs of $\alpha'$ values estimated for the first and last authors for the same journals, we found that $\alpha'$ tended to be higher for first authors than for last authors (Figure S4; Effect of the fixed factor 'Authorship position' in a linear mixed model: Cohen's $d$ = $0.065{\pm}0.02$, $t_{2024}$ = 4.28, p < 0.0001). This suggests that the gender of the first author was a slightly stronger predictor of the remaining authors' genders than the gender of the last author, i.e. the opposite of what is predicted if senior scientists are causally responsible for homophily.

## Variance in homophily between disciplines
\autoref{alpha_histograms} illustrates the variance in journal homophily values ($\alpha'$) across scientific disciplines. All disciplines had a positive average $\alpha'$, although homophily appeared somewhat stronger in some disciplines than others (e.g. mean $\alpha'$ was $0.12{\pm}0.02$ for Urology journals and $0.03{\pm}0.01$ for Veterinary Medicine journals; \autoref{alpha_histograms}, Table S3). However, there was no evidence for consistent differences in $\alpha'$ between disciplines: the random factor 'Discipline' explained around 1% of the variance in $\alpha'$ in the two linear mixed models described in the previous section (see \autoref{alpha_histograms} and mixed models in Online Supplementary Material). Thus, we cannot reject the null hypothesis that the processes causing positive $\alpha'$ are similarly strong in all the disciplines we examined.

There was no indication that journals publishing on a wide range of topics have higher $\alpha'$ values than more specialised journals, due to the Wahlund effect. For example, the journal category 'Multidisciplinary' -- which includes journals like _PLoS ONE_, _Nature_, _Science_, and _PNAS_ -- did not have notably elevated $\alpha'$ (\autoref{alpha_histograms}). This result suggests that our estimates of homophily, and estimates from some earlier studies, are not notably inflated by the presence of disparate research topics (with variable author gender ratios) being published within individual journals.

@@

## Analysis accounting for differences in author gender ratio between countries
When we restricted the analysis by country, we observed statistically significant homophily for 72 of the 325 journal-country combinations tested (64 unique journals and 18 unique countries), and no significant evidence of heterophily (Figures S5-S6). Additionally, the values of $\alpha'$ calculated for each journal-country combination were mostly very similar to the $\alpha'$ values calculated for the journal as a whole (i.e. when pooling papers from different countries); on average, $\alpha'$ was only inflated in the main (non-country specific) analysis by 0.002 (Figure S7). These results suggest that our findings of widespread homophily in the main analysis were not driven solely by a Wahlund effect resulting from gender differences between countries.

@@

![When there is a difference in gender ratio between early-career and established researchers, and collaboration is non-random with respect to career stage, the null expectation for $\alpha$ deviates from zero. An excess of collaborations between career stages gives the appearance of gender heterophily (lower rows, blue areas), while an excess of within-career stage collaborations produced apparent gender homophily (upper rows, red areas). However, the conditions required for strong gender homophily are quite restrictive, making it unlikely that this issue explains all of the homophily observed in Figure 2. Contour lines denote increments of 0.1. \label{simulation}](../figures/figure5 - Simulation.pdf)

@@

# Discussion
We found evidence that researchers preferentially publish with same-gendered coauthors, even after implementing stringent controls for Wahlund effects (\autoref{wahlund_plot}). Our study therefore reaffirms earlier studies' conclusions [e.g. @Ferber_1980; @McDowell_1992; @Bentley_2003; @Crow_2015; @Ghiasi_2015; @Araujo_2017a; @Fahmy_2017; @Jadidi_2017; @Teele_2017; @Zettler_2017] and establishes their generality across the life sciences. Relatively few journals had $\alpha'$ values below zero, and almost no journals showed statistically significant gender heterophily after controlling for multiple testing. The excess of same-gender coauthorships was quite large: many journals had $\alpha' > 0.1$, indicating that the gender ratio of men's and women's coauthors differs by >10% in absolute terms. In relative terms, the observed homophily is even more striking: for example, if the proportion of male coauthors is 80% for men and 70% for women ($\alpha' = 0.1$), the results imply that women publish with women 50% more frequently than men do.

@@

Some disciplines exhibit especially skewed gender ratios; e.g. most authors in physics and computer science are men, while most authors in nursing are women [@Holman_2018]. We hypothesised that these gender-imbalanced disciplines might show the strongest gender homophily, e.g. because being in the minority might increase one's drive to seek out same-gendered colleagues. Contrary to this hypothesis, we found no evidence that gender homophily is restricted to particular disciplines: $\alpha'$ was similarly high across the board (\autoref{alpha_histograms}). Moreover, $\alpha'$ was lower for journals with a male-biased author gender ratio relative to those with an even gender ratio. Interestingly, $\alpha'$ was highest in journals with a female-biased author gender ratio. One interpretation of this result is that men are more likely to preferentially seek out male collaborators in fields where men are a minority, relative to the homophily displayed by women in fields where women are a minority. However, this latter result is only tentatively supported since there are so few journals in which most authors are women (see \autoref{alpha_gender_ratio}).

We also found that gender homophily was marginally stronger in 2015-2016 relative to 2005-2006. Although this trend might reflect a change in the gender preferences of researchers seeking collaborators over time, there are alternative (and perhaps more likely) explanations. For example, this trend might result from the increasing number of women working in senior positions in STEMM over the past decade [e.g. @Long_2015; @Bendels_2018; @McKenzie_2017]. As shown in \autoref{simulation}, if enough coauthorships are between junior and senior researchers, a large gender gap between career stages can give the appearance of heterophily. As this gender gap between career stages lessens, the observed values of $\alpha$ may increase.

Our study begs two questions: what causes gender homophily in science, and are our results cause for concern? These questions are closely related. For example, some of the homophily we observed might be caused by women seeking to avoid harassment, sexism or microaggressions from men [@jagsi_2016], which would clearly be concerning. Additionally, Sheltzer and Smith [-@sheltzer_2014] concluded that 'elite' male academics (defined as recipients of major honours) have a higher proportion of male students and postdocs than non-elite male academics. This finding could contribute to the homophily we observed, and is cause for concern since Sheltzer and Smith [-@sheltzer_2014]'s results might reflect discrimination against women during hiring [@Moss_2012], or avoidance by women of elite research groups (e.g. due to 'imposter syndrome' or the perception that these groups are sexist). We also found a little evidence that gender homophily is detrimental to research quality, in that high-impact journals tended to have weaker homophily. Assuming that papers published in high-impact journals are of higher average quality [which is contentious; @Garfield_2006], this result provides non-experimental support for the hypothesis that mixed-gender teams produce better research than single-gender teams [@Campbell_2013].

Homophily might also have more benign causes. Collaboration is often most enjoyable and productive when working with like-minded people, who might be same-gendered more often than not. We also suppose that some people consciously choose to preferentially collaborate with women in order to help close the gender gap in the workforce; this would create homophily if women do this more than men. In support of this interpretation, women appear more likely than men to promote the work of female collagues by inviting them to give talks [@nittrouer_2018; @debarre_2018]. Related to this, Ghiasi et al. [-@Ghiasi_2015] concluded that women in engineering are "compliant [in reproducing] male-dominated scientific structures" because they do not collaborate often enough with other women; their data suggest that coauthorships between two women are about 30% more frequent than expected under the null (see their Figure 7). In contrast, we propose that it behoves men as well as women to ensure that they are not overlooking or excluding female colleagues. Given that women work disproportionately often in junior positions with little opportunity to be selective, one could argue that the responsibilty to address gender imblances in science rests primarily with men.

@@

Although the proportion of women in science has significantly increased over the last decades, women still tend to leave their academic careers behind considerably more often than their male counterparts [REF]. Lack of scientific recognition [@Lincoln_2012], lack of research funding [@Bozeman_2004], lack of invitation to speak at scientific events [@Isbell_2012; @Klein_2017; @debarre_2018], discrimination during hiring [@MossRacusin_2017; @Reuben_2014] and discrimination when writing a recommendation letter [@Trix_2003; @Schmader_2007], are only some of the small disadvantages adding up to drive women out of science. Whether their decision to leave academia is deliberate or not, it will negatively impact science overall, as mentoring, networking, and supporting younger women scientists are especially important for women in academia.
